

---
title: "Exam 1: US President Inauguration Speech Length and Party Affiliation"
author: "Kyle Ke"
date: "10/27/2021"
abstract: "The analysis investigates whether there is a statistically significant difference in the inauguration speech word count between US presidents who are Republican vs Democrats. It is performed using a two-sample t-test, Mann–Whitney U test, approximate permutation test, and bootstrap test. The results conclude that there is no statistically significant difference in inauguration speech length between Republican and Democratic presidents."
output:
  pdf_document:
    toc: yes
  word_document:
    toc: yes
  html_document:
    toc: yes
    df_print: paged
bibliography: 
- report.bib
- packages.bib
biblio-style: apalike
link-citations: yes
nocite: '@*'
header-includes:
  \usepackage{placeins}
  
  
  
---




```{r,echo=FALSE,results = FALSE,message=FALSE}
knitr::opts_chunk$set(message = FALSE)
```



```{r,echo=FALSE,results = FALSE}
library(rvest)
library(dplyr)
library(ggplot2)
library(fBasics)
library(knitr)
library(formatR)
```

```{r bib, include=FALSE}
knitr::write_bib(c('rvest','dplyr','ggplot2','fBasics','knitr','formatR'),
                 file='packages.bib',width = 60)
```


```{r,echo=FALSE,results = FALSE}
length_of_inaugural_url = "https://www.potus.com/presidential-facts/inaugural-address-length/"
length_of_inaugural_list = rvest::read_html(length_of_inaugural_url)
length_of_inaugural_df = length_of_inaugural_list %>% rvest::html_table() %>% as.data.frame(.)

party_affilation_url = "https://www.potus.com/presidential-facts/political-party/"
party_affilation_list = rvest::read_html(party_affilation_url)
party_affilation_df = party_affilation_list %>% rvest::html_table() %>% as.data.frame(.)
```


```{r,echo=FALSE,results = FALSE}
length_of_inaugural_df = length_of_inaugural_df[-c(60:64),]
party_affilation_df = party_affilation_df[-c(1,21,38:51),]
party_affilation_df$Party = c(rep('Republican',19),rep('Democrat',16))

```


```{r,echo=FALSE,results = FALSE}
joint_df = merge(party_affilation_df,length_of_inaugural_df,by="President")
joint_df$Length.in.Words = as.numeric(gsub(",", "", joint_df$Length.in.Words))
joint_df = joint_df[,-c(2,3,4,6,7,8)]
```



\newpage



# Introduction

The majority (35 out of 47) of all US presidents, are affiliated with either the Republican or the Democratic Party. Since the inauguration speech word count of all US presidents are available, it would be interesting to see if there is a difference in inauguration speech length between Republican presidents and Democratic presidents.


## Methdology

The inauguration speech word count of all 24 inauguration speeches by Republican presidents and 23 inauguration speeches by Democratic presidents were collected from the website by (@potus). We have 24 observations, denoted $x_1, x_2, . . . x_{24}$ in the Democratic group, and 23 observations, denoted $y_1, y_2, . . . y_{23}$ in the Republican group. A two-sample t-test, Mann-Whitney U test, approximate permutation test, and bootstrap test were performed during the analysis. 

The approximate permutation test (@helwig_2017) was choosen because the number of combinations for  ${47 \choose 24}$ is 1.612 $\times$ $10^{13}$. The number of all combinations is too big for us to compute all of the test statistics (difference in means) for each combination due to limited computing resources. Instead, a random sample of 24 word counts was taken 10000 times. Each sample of 24 was taken without replacement. However, the same combination could theoretically appear in more than one sample because we could not enumerate all combinations. The mean of each sample was taken and subtracted from the mean of its complement (i.e. the other 23 word counts) to obtain our test statistic: difference in mean. All 10000 test statistics were then pooled as a single distribution. The p-value was obtained by counting all of the absolute value of the test statistics that are greater than the absolute value of the original test statistic and dividing that count by 10000.

A bootstrap (randomization) test was performed taking a random sample of 47 word counts 10000 times. However, each sample of 47 word counts was taken with replacement, meaning the same word count could appear more than once within a sample. Within each sample, the mean of the word counts indexed [1:24] was subtracted from the mean of the word counts indexed [25:47] to obtain our test-statistic. All 10000 test statistics were then pooled as a single distribution. The p-value was obtained by counting all of the absolute value of the test statistics that are greater than the the absolute value of the original test statistic and dividing that count by 10000.






The null and alternative hypothesis are as follows:

*Two-sample t-test, Approximate Permutation Test, and Bootstrap Test:* (@helwig_2017)

$H_o: \mu_x = \mu_y$ The mean inauguration speech word count is the same between Democratic and Republican presidents 

$H_A: \mu_x \neq \mu_y$ The mean inauguration speech word count is different between Democratic and Republican presidents

Significance Level: .05

*Mann-Whitney U test:*(@shier)

$H_o: P(x_i>y_j) = .5$ The median inauguration speech word count is the same between all Democratic and Republican presidents

$H_A: P(x_i>y_j) \neq .5$ The median inauguration speech word count is not the same between all Democratic and Republican Presidents


Significance Level: .05

\newpage
# Data Description

*Summary Statistics*

```{r, echo=FALSE}
Republicans=joint_df[joint_df$Party=="Republican",]$Length.in.Words
Democrats = joint_df[joint_df$Party=="Democrat",]$Length.in.Words
options(scipen = 0,digits=6)
summary_stats = cbind(basicStats(Republicans),basicStats(Democrats))
summary_stats = summary_stats[c("Minimum","Maximum","1. Quartile","3. Quartile",
                                "Mean","Median","Stdev"),] 
knitr::kable(summary_stats,caption = "Summary Statistics for Inaugration Speech Word Counts",
             col.names=colnames(summary_stats))

```





*Box plots*

```{r,echo=FALSE,boxplot,fig.cap = "Inauguration Speech Word Count Boxplots by Party",fig.width=7, fig.height=5,fig.align='Center'}

ggplot(joint_df, aes(x=as.factor(Party), y=Length.in.Words,fill=Party)) + 
    geom_boxplot(alpha=0.2) + 
    ggtitle('                 Inauguration Speech Word Count by Party ') +
ylab('Word Counts') + xlab('Party')
```


We observe from the box plots and summary statistics that the median word count for Republican presidents is higher than the median word count for Democratic presidents. Moreover, the variance in word count is higher for Republican president compared to Democratic presidents.

*Quantile-Quantile Plots*

```{r,echo=FALSE,fig.cap = "QQ Plot of Inaguration Speech Word Count for Democratic Presidents",fig.width=5, fig.height=3,fig.align='Center'}
ggplot(joint_df[joint_df$Party=="Democrat",],aes(sample = Length.in.Words )) +
  stat_qq() +
  stat_qq_line()  + ggtitle('            QQ Plot for Democratic Presidents ') +
xlab('Theoretical Quantile') + ylab('Observed Word Counts')
```

\newpage
```{r,echo=FALSE,fig.cap = "QQ Plot of Inaguration Speech Word Count for Republican Presidents",fig.width=5, fig.height=3,fig.align='Center'}

ggplot(joint_df[joint_df$Party=="Republican",],aes(sample = Length.in.Words )) +
  stat_qq() +
  stat_qq_line()  + ggtitle('            QQ Plot for Republican Presidents ') +
xlab('Theoretical Quantile') + ylab('Observed Word Counts')

```


The quantile-quantile plots show that inauguration speech word counts for Democratic presidents is heavily right skewed. Thus, the data is exhibiting properties that might be more suitable for the non-parametric Mann-Whitney U test.

\newpage

# Results

*Two sample t-test*

```{r,echo=FALSE,results = FALSE}
t_Test_results <- t.test(Republicans,Democrats,"two.sided" ,paired = FALSE,var.equal = TRUE,fig.align='Center')
t_Test_results
```
\vspace{12pt}

The two sample t-test has a p-value of `r round(t_Test_results$p.value,digits=3)`. Under a significance level of .05, we fail to reject the null hypothesis that the mean inauguration speech word count is the same between Democratic and Republican presidents.



\vspace{12pt}



*Mann-Whitney U test*
```{r,echo=FALSE,results = FALSE,fig.align='Center'}
wilcox_test_results <- wilcox.test(Republicans,Democrats, exact = TRUE, correct = TRUE, 
                                   alternative = 'two.sided')
```
\vspace{12pt}


The (Mann-Whitney U test has a p-value of `r round(wilcox_test_results$p.value,digits=3)`. Under a significance level of .05, we fail to reject the null hypothesis that the median inauguration speech word count is the same between Democratic and Republican presidents.

\vspace{12pt}


The results from the t-test and Man-Whitney U test are in congruent with each other. At the .05 significance level, both tests fail to reject the null hypothesis that there is a difference in the central tendency of inauguration speech length between Democratic and Republican presidents. 



*Approximate Permutation Test*
```{r,echo=FALSE,fig.cap = "Histogram of Difference in Means from Approx Permutation Tests",fig.width=5, fig.height=3,fig.align='Center'}
set.seed(5)
meandif <- function(x,y) mean(x) - mean(y)
perm2samp <- function(x,y,myfun=meandif,nsamp=10000,
                      alternative=c("two.sided","less","greater")){
  theta.hat = myfun(x,y)
  m = length(x)
  n = length(y)
  N = m + n
  z = c(x,y)
  gmat = replicate(nsamp,sample.int(N,m))
  theta.mc = apply(gmat,2,function(c,z){myfun(z[c],z[-c])},z=z)
  if(alternative[1]=="less"){
    aslperm = sum(theta.mc <= theta.hat) / nsamp
  } else if(alternative[1]=="greater"){
    aslperm = sum(theta.mc >= theta.hat) / nsamp
  } else{
    aslperm = sum(abs(theta.mc) >= abs(theta.hat)) / nsamp
  }
  list(theta.hat=theta.hat,theta.mc=theta.mc,asl=aslperm)
  return(list(theta.hat,theta.mc,aslperm))
}


perm_results = perm2samp(Republicans,Democrats,myfun=meandif,nsamp=10000,alternative="two.sided")
theta.sampled=perm_results[[2]]
ggplot() +aes(theta.sampled)+ geom_histogram() + ggtitle('           Difference in Means (Approx Perm) ') + xlab('Difference In Mean') +  geom_vline(xintercept = perm_results[[1]], color = "blue", size=1.5)

```
The results of the approximate permutation test yielded a p-value (two-sided) of `r round(perm_results[[3]],digits=3)`. Under the significance level of .05, we fail to reject the null hypothesis that the mean inauguration speech word count between Democratic and Republican presidents is the same.

\newpage


*Bootstrap Test*
```{r,echo=FALSE,fig.cap = "Histogram of Difference in Means from Bootstrap",fig.width=5, fig.height=3,fig.align='Center'}
set.seed(5)
meandif <- function(x,y) mean(x) - mean(y)
perm2samp <- function(x,y,myfun=meandif,nsamp=10000,
                      alternative=c("two.sided","less","greater")){
  theta.hat = myfun(x,y)
  m = length(x)
  n = length(y)
  N = m + n
  z = c(x,y)
  gmat = replicate(nsamp,sample.int(N,N,replace=TRUE))
  theta.mc = apply(gmat,2,function(c,z){myfun(z[c[1:23]],z[c[24:47]])},z=z)
  if(alternative[1]=="less"){
    aslperm = sum(theta.mc <= theta.hat) / nsamp
  } else if(alternative[1]=="greater"){
    aslperm = sum(theta.mc >= theta.hat) / nsamp
  } else{
    aslperm = sum(abs(theta.mc) >= abs(theta.hat)) / nsamp
  }
  list(theta.hat=theta.hat,theta.mc=theta.mc,asl=aslperm)
  return(list(theta.hat,theta.mc,aslperm))
}


perm_results = perm2samp(Republicans,Democrats,myfun=meandif,nsamp=10000,alternative="two.sided")
theta.sampled=perm_results[[2]]
ggplot() +aes(theta.sampled)+ geom_histogram() + ggtitle('           Difference in Means (Bootstrap) ') + 
  xlab('Difference In Mean') +  geom_vline(xintercept = perm_results[[1]], color = "blue", size=1.5)

```
The results of the bootstrap test yielded a p-value (two-sided) of `r round(perm_results[[3]],digits=3)`. Under the significance level of .05, we fail to reject the null hypothesis that the mean inauguration speech word count between Democratic and Republican presidents is the same.




# Conclusions

Although, through eye observations of the summary statistics and box plots, Republican presidents appear to have generally longer inauguration speeches, four different tests have failed to detect a statistically significant difference in inauguration speech word count between Democratic presidents and Republican presidents.

# Recommedations for Future Studies

Future studies could investigate the two outliers in the Democratic president group as shown in Figure 1. It is possible the results of the analysis could change as the t-test is especially sensitive to outliers. Future studies could also consider conducting one-tailed tests to determine if the speech length of Republican presidents is greater than the speech length of Democratic presidents, instead of only investigating if there is any difference between the groups as done in this analysis. Moreover, future studies could investigate if there is difference across time (e.g., 1800s, 1900s, and 2000s).

\newpage

# References
::: {#refs}
:::

\newpage

# Appendix

## Code
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=100), tidy=TRUE)
```





